---
layout: page
title: dplyr
reading: "<a href='https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html'>Vignette</a>, <a href='https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf'>Data wrangling cheatsheet</a>"
notes: "Screencast - dplyr <a href='https://www.youtube.com/watch?v=pPWDD34o4mE'>part 1</a>, <a href='https://www.youtube.com/watch?v=LRF7nS4xdu4'>part 2</a>"
output: 
    ioslides_presentation:
        widescreen: true
slides: true
link: true
---

```{r, echo=FALSE}
options(width = 110)
```

# dplyr

## A Grammar of Data Manipulation {.smaller}

dplyr is based on the concepts of functions as verbs that manipulate data frames.

<br/>

Single data frame functions / verbs:

* `tbl_df()`: add the `tbl_df` class 
* `filter()`: pick rows matching criteria
* `slice()`: pick rows using index(es)
* `select()`: pick columns by name
* `rename()`: rename specific columns
* `arrange()`: reorder rows
* `mutate()`: add new variables
* `transmute()`: create new data frame with variables
* `distinct()`: filter for unique rows
* `sample_n()` / `sample_frac()`: randomly sample rows
* `summarise()`: reduce variables to values


## dplyr rules for functions

* First argument is a data frame

* Subsequent arguments say what to do with data frame

* Always return a data frame

* Don't modify in place


## Example Data {.smaller}

We will demonstrate dplyr's functionality using the nycflights13 data.

```{r}
suppressMessages(library(dplyr))
library(nycflights13)

flights = flights %>% tbl_df()
str(flights)
```

## tbl_df()? {.smaller}
```{r}
flights
```

## filter() - March flights {.smaller}

```{r}
flights %>% filter(month == 3)
```

## filter() - Flights in the first 7 days of March {.smaller}

```{r}
flights %>% filter(month == 3, day <= 7)
```

## filter() - Flights to LAX or RDU in March {.smaller}

```{r}
flights %>% filter(dest == "LAX" | dest == "RDU", month==3)
```

## slice() - First 10 flights {.smaller}

```{r}
flights %>% slice(1:10)
```

## slice() - Last 5 flights {.smaller}

```{r}
flights %>% slice((n()-5):n())
```

## select() - Individual Columns {.smaller}

```{r}
flights %>% select(year, month, day)
```

## select() - Exclude Columns {.smaller}

```{r}
flights %>% select(-year, -month, -day)
```

## select() - Ranges {.smaller}

```{r}
flights %>% select(year:day)
```

## select() - Exclusion Ranges {.smaller}

```{r}
flights %>% select(-(year:day))
```

## select() - Matching {.smaller}

<div class="columns-2">
```{r}
flights %>% select(contains("dep"), 
                   contains("arr"))
```
```{r}
flights %>% select(contains("dep"), 
                   starts_with("arr"))
```
</div>


## rename() {.smaller}

```{r}
flights %>% rename(tail_number = tailnum)
```

## arrange() {.smaller}

```{r}
flights %>% filter(month==3,day==2) %>% arrange(origin, dest)
```

## arrange() & desc() {.smaller}

```{r}
flights %>% filter(month==3,day==2) %>% arrange(desc(origin), dest)
```

## mutate() {.smaller}

```{r}
library(lubridate)
flights %>% select(1:3) %>% mutate(date = paste(month,day,year,sep="/") %>% mdy())
```

## transmute() {.smaller}

```{r}
flights %>% select(1:3) %>% transmute(date = paste(month,day,year,sep="/") %>% mdy())
```

## distinct() {.smaller}

```{r}
flights %>% select(origin, dest) %>% distinct() %>% arrange(origin,dest)
```

## sample_n() {.smaller}

```{r}
flights %>% sample_n(10)
```

## sample_frac() {.smaller}

```{r}
flights %>% sample_frac(0.001)
```

## summarise() {.smaller}

```{r}
flights %>% mutate(date = paste(month,day,year,sep="/") %>% mdy()) %>% 
            summarize(n(), min(date), max(date))
```

## group_by() {.smaller}

```{r}
flights %>% group_by(origin)
```

## summarise() with group_by() {.smaller}

```{r}
flights %>% group_by(origin) %>%
            mutate(date = paste(month,day,year,sep="/") %>% mdy()) %>% 
            summarize(n(), min(date), max(date))
```

## Exercises

1. Which plane (check the tail number) flew out of each New York airport the most?

1. What was the shortest flight out of each airport in terms of distance? In terms of duration?

1. How many flights to Los Angeles (LAX) did each of the legacy carriers (AA, UA, DL or US) have in May from JFK, and what was their average duration?

1. Which date should you fly on if you want to have the lowest possible average depature delay? What about arrival delay?

1. Create a time series plot of each of the legacy carriers' average departure delay by day and origin airport.



